---
title: "Analysis of World's Top Intelligent Individuals"
author: "Hasan Sezer"
date: "2024-08-16"
output:
  
  pdf_document: default
  html_document: default
---


## 1. Data Preparation

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
```

```{r}
topIntelligent <- read.csv("top_intelligent.csv")
glimpse(topIntelligent)
```
## 2. Data Cleaning
Check for "N/A" values
```{r}
sapply(topIntelligent, function(x) sum(x == "N/A"))
```


```{r}
table(topIntelligent$Awards)
```
Awards column has 1249 NAs in string format. 

Drop NAs in Awards

```{r}
topIntelligent_cleaned <- subset(topIntelligent, Awards != "N/A")
table(topIntelligent_cleaned$Awards)
```

## 3. Descriptive Statistics
### 3.1 Distribution of Awards by Each Country


```{r}
Awards_by_Country <- topIntelligent_cleaned %>%
  count(Country, Awards)
```

```{r}
ggplot(data = Awards_by_Country, aes(x = Country, y = n, fill = Awards)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Number of Awards by Country",
       x = "Country",
       y = "Count") +
  theme_minimal()
```

### 3.2 The Country with the Most Nobel Prizes

```{r}
NobelPrize <- Awards_by_Country %>%
  filter(Awards == "Nobel Prize")

```

```{r}
ggplot(data = NobelPrize, aes(x = Country, y = n)) +
  geom_bar(stat = "identity", position = "dodge", fill = "Skyblue") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Number of Nobel Prizes by Country",
       x = "Country",
       y = "Count") +
  theme_minimal()
```

### 3.3 The Country with the Most Awards

```{r}
sum_country_awards <- topIntelligent %>%
  group_by(Country) %>%
  summarize(Total_Awards = n())

sum_country_awards %>%
  arrange(desc(Total_Awards))
```

```{r}
ggplot(data = sum_country_awards, aes(x = Country, y = Total_Awards, fill = Country)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Total Awards by Country",
       x = "Country",
       y = "Count") +
  theme_minimal()
```

### 3.4 IQ Analysis

```{r}
summary(topIntelligent$IQ)
```

IQ and Country
```{r}
IQ_by_Country <- topIntelligent %>%
  group_by(Country) %>%
  summarize(Average_IQ = mean(IQ, na.rm = TRUE))

IQ_by_Country

```

```{r}
ggplot(data = IQ_by_Country, aes(x = Country, y = Average_IQ, fill = Country)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Average IQ by Country",
       x = "Country",
       y = "Average IQ") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**Interpretation:**

There is almost no IQ variance by country. 

- IQ and Education
```{r}
table(topIntelligent$Education)
```

```{r}
IQ_by_Education <- topIntelligent %>%
  group_by(Education) %>%
  summarise(Avr_IQ = mean(IQ, na.rm = TRUE))

IQ_by_Education
```
- IQ and Influence

```{r}
table(topIntelligent$Influence)

```

```{r}
IQ_by_Influence <- topIntelligent %>%
  group_by(Influence) %>%
  summarize(Average_IQ = mean(IQ, na.rm = TRUE))

IQ_by_Influence

```


- IQ by Awards
```{r}
IQ_by_Awards <- topIntelligent_cleaned %>%
  group_by(Awards) %>%
  summarise(Avr_IQ = mean(IQ, na.rm = TRUE))

IQ_by_Awards
```

**Interpretation:**
IQ has not shown variance across the variables inspected so far. 

### 3.5 Fields Analysis

- Field with the most awards

```{r}

award_by_field <- topIntelligent_cleaned %>%
  count(Field.of.Expertise, Awards)
```


```{r}

field_with_most_awards <- award_by_field %>%
  group_by(Field.of.Expertise) %>%
  summarise(total_awards = sum(n))
```


```{r}
ggplot(data = field_with_most_awards, aes(x = Field.of.Expertise, y = total_awards)) +
  geom_bar(stat = "identity", position = "dodge", fill = "orange") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Total Awards by Fields",
       x = "Field of Expertise",
       y = "Count") +
  theme_minimal()
```

- Distribution of different awards in varying fields. 

```{r}
field_by_award <- topIntelligent_cleaned %>%
  count(Field.of.Expertise, Awards)

ggplot(data = field_by_award, aes(x = Field.of.Expertise, y = n, fill = Awards)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Total Awards by Fields",
       x = "Field of Expertise",
       y = "Count") +
  theme_minimal()
```

**Interpretation:**
Physics has the most awards. In Physics, the Nobel Prize is granted to most. 

### 4. How Can One Get a Nobel Prize: Regression Analysis

In this section, I will try to predict what factors contribute to receiving a Nobel Prize.

### 4.1 Logistic Regression Analysis

### Creating a New Column for the Nobel Prize

I will create a new column, `NobelPrize,` with dummy variables:
- `0` will be the reference category for not getting a Nobel Prize.
- `1` will indicate receiving a Nobel Prize.

```{r create-dummy-variable}
# Create the Nobel Prize dummy variable
topIntelligent_cleaned$NobelPrize <- ifelse(topIntelligent_cleaned$Awards == "Nobel Prize", 1, 0)
```

```{r}
str(topIntelligent_cleaned)
```
```{r}
table(topIntelligent_cleaned$NobelPrize)
```

```{r}
table(topIntelligent_cleaned$NobelPrize, topIntelligent_cleaned$Country)
```

I must convert categorical variables from string format to factors for the logistic regression model.

```{r}
topIntelligent_cleaned$Country <- as.factor(topIntelligent_cleaned$Country)
topIntelligent_cleaned$Field.of.Expertise <- as.factor(topIntelligent_cleaned$Field.of.Expertise)
topIntelligent_cleaned$Achievements <- as.factor(topIntelligent_cleaned$Achievements)
topIntelligent_cleaned$Gender <- as.factor(topIntelligent_cleaned$Gender)
topIntelligent_cleaned$Education <- as.factor(topIntelligent_cleaned$Education)
topIntelligent_cleaned$Influence <- as.factor(topIntelligent_cleaned$Influence)
```


#### Evaluate Model Fit Using Backward Selection

I will evaluate the model fit using a backward selection method. I include all variables in the model and remove them individually, checking the model fit using AIC (Akaike Information Criterion) and Deviance.


```{r}
backward_model <- glm(NobelPrize ~ Country + Field.of.Expertise + Achievements + Gender + Education + Influence + IQ, data = topIntelligent_cleaned, family = binomial())

stepwise_backward_model <- step(backward_model, direction = "backward")

```


```{r}
# summary(backward_model)
```

**Interpretation:**
Only IQ turns out to be significant in predicting Nobel Prize wins. Removing this variable degrades the model fit (i.e., increases AIC).


- Validating significant predictor 


```{r}
# Confidence intervals for the model coefficients
confint.default(backward_model)

# Exponentiated coefficients
exp(coef(backward_model))

library(broom)
tidy(stepwise_backward_model, exponentiate = TRUE)
```


-- To validate regression model output, now I will use the Chi-Squared test to find the association between the outcome and some categorical variables at a time.


```{r}
chisq.test(table(topIntelligent_cleaned$Country, topIntelligent_cleaned$NobelPrize))
```

```{r}
chisq.test(table(topIntelligent_cleaned$Gender, topIntelligent_cleaned$NobelPrize))
```
```{r}
chisq.test(table(topIntelligent_cleaned$Education, topIntelligent_cleaned$NobelPrize))
```

```{r}
chisq.test(table(topIntelligent_cleaned$Field.of.Expertise, topIntelligent_cleaned$NobelPrize))
```

```{r}
chisq.test(table(topIntelligent_cleaned$Achievements, topIntelligent_cleaned$NobelPrize))
```
**Interpretation:**
Again, no categorical variables under investigation are significant in predicting getting a Nobel Prize. 

### 4.1 Lasso Regression Analysis

I will try out Lasso Regression to find which variables effectively predict a Nobel Prize win.  

```{r}
# install.packages("glmnet")
```

```{r}
library(glmnet)
```

```{r}
data_dummy <- model.matrix(NobelPrize ~ Country + Field.of.Expertise + Achievements + Gender + Education + Influence + IQ - 1, data = topIntelligent_cleaned)
```

```{r}
X <- data_dummy
y <- topIntelligent_cleaned$NobelPrize
```

```{r}
lasso_model <- cv.glmnet(X, y, family = "binomial", alpha = 1)
```

```{r}
best_lambda <- lasso_model$lambda.min
coef(lasso_model, s = best_lambda)
```

**Interpretation:**

The Lasso Regression model finds the most relevant predictors to the outcome variable. Here, non-zero coefficients contribute to predicting the Nobel Prize win. Only the USA and IQ have non-zero coefficients. Also, IQ slightly affects the Nobel Prize win: Higher IQ is associated with a slight increase in the log-odds of receiving a Nobel Prize. On the other hand, being a US citizen decreases receiving a Nobel Prize, which the following contingency table can confirm. 

```{r}
table(topIntelligent_cleaned$Country, topIntelligent_cleaned$NobelPrize)
```

**Summary:**

In this project, I have worked on the Top Intelligent People in the world and tried to gain insight into their profiles using EDA and regression analysis. EDA has provided descriptive information on questions regarding single and multi-variables. Some include the Distribution of Awards by Country, the Distribution of Awards by Country, IQ and its relation with variables (e.g., Awards, Education, Country, etc.), and the Distribution of Awards by Fields. Questions in EDA can be expanded depending on the needs. 

In the last section, I wondered what predictors may contribute to winning a Nobel Prize. The results from Logistic Regression showed that only IQ is significant in predicting winning a Nobel Prize. 

Further, I wanted to validate the results of Logistic Regression by building a Lasso Regression model, which helps identify which predictors (variables) are most important by shrinking the less significant coefficients to zero. This model is helpful in datasets with many multicollinear variables, as it can help simplify the model by focusing on the most relevant features/variables. With Lasso Regression, I wanted to increase the predictive accuracy of my model by selecting essential variables and better understanding the key factors influencing Nobel Prize achievements.

The results of the Lasso Regression model revealed that IQ is slightly influential in predicting Nobel Prize wins, while the USA is negatively associated with winning a Nobel Prize.  









